Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-07-31 paper.qmd
Updated Figure 8

## To obtain the standardise counts within hexbins
counts_df1 <- hb_obj_scurve1$std_cts
df_bin_centroids1 <- extract_hexbin_centroids(
  centroids_df = all_centroids_df1, 
  counts_df = counts_df1) |>
  filter(drop_empty == FALSE) |>
  mutate(a1 = a1_1)

@@

## To obtain the standardise counts within hexbins
counts_df2 <- hb_obj_scurve2$std_cts
df_bin_centroids2 <- extract_hexbin_centroids(
  centroids_df = all_centroids_df2, 
  counts_df = counts_df2) |>
  filter(drop_empty == FALSE) |>
  mutate(a1 = a1_2)

@@

## To obtain the standardise counts within hexbins
counts_df3 <- hb_obj_scurve3$std_cts
df_bin_centroids3 <- extract_hexbin_centroids(
  centroids_df = all_centroids_df3, 
  counts_df = counts_df3) |>
  filter(drop_empty == FALSE) |>
  mutate(a1 = a1_3)

@@

The default number of bins $b=b_1\times b_2$ is computed based on the sample size, by setting $b_1=n^{1/3}$, consistent with the Diaconis-Freedman rule [@freedman1981]. The value of $b_2$ is determined analytically by $b_1, q, r_2$. Values of $b_1$ between $2$ and $b_1 = \frac{n}{2}$ are allowed. FIG XXX(b) shows the effect of different choices of $b_1$ on the MSE of the fitted model. XXX Are you showing $b_1$ or $b$ here???? $b$

@@

<!--To generate errors for different total number of bins-->
```{r}
#| label: errors-scurve
## To initialize number of bins along the x-axis
bin1_vec_scurve <- 2:19 #sqrt(NROW(training_data_scurve)/2)
buffer_vec <- seq(0.05, 0.1, by = 0.01)

@@

for (xbins in bin1_vec_scurve) {
  for (buff in buffer_vec) {

    bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = buff)$bin2

    scurve_model <- fit_highd_model(
      training_data = training_data_scurve,
      emb_df = umap_scurve_scaled,
      bin1 = xbins,
      r2 = r2,
      q = buff, 
      is_bin_centroid = TRUE,
      is_rm_lwd_hex = FALSE,
      col_start_highd = "x"
    )

@@

    ## Compute error
    error_df <- glance(
      df_bin_centroids = df_bin_centroids_scurve,
      df_bin = df_bin_scurve,
      training_data = training_data_scurve,
      newdata = NULL,
      type_NLDR = "UMAP",
      col_start = "x") |>
      mutate(bin1 = xbins,
             bin2 = bin2,
             b = bin1 * bin2,
             b_non_empty = NROW(df_bin_centroids_scurve),
             buff = buff)

@@

@@

mse_scurve_b <- ggplot(error_scurve, 
                     aes(x = b, 
                         y = MSE,
                         color = as.factor(buff),
                         group = as.factor(buff))) + 
  geom_point(size = 1) +
  geom_line(linewidth = 0.05) + 
  # geom_vline(xintercept = 220, linetype="solid", 
  #            color = "black", linewidth=0.1) +
  # geom_vline(xintercept = 112, linetype=2, 
  #            color = "black", linewidth=0.1) +
  # geom_vline(xintercept = 364, linetype=2, 
  #            color = "black", linewidth=0.1) +
  scale_colour_discrete_qualitative(palette = "Dark 2") +
  labs(x = expression(b), y = "MSE", color = expression(q)) +
  ggtitle("(b)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12, 
            angle=90),
        plot.title = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        legend.key.height = unit(1, 'cm'),
        legend.key.width = unit(1, 'cm'),
        legend.key.size = unit(1, 'cm'))
```

@@

cell_count_scurve <- ggplot(df_bin_centroids_all, 
                            aes(x = reorder(as.factor(hexID), 
                                            -bin_counts), 
                                y = bin_counts,
                                group = as.factor(a1),
                                color = as.factor(a1))) + 
  geom_line() + 
  # geom_hline(yintercept = benchmark1, color = "black", linewidth=0.5, alpha = 0.5) +
  scale_colour_discrete_qualitative(palette = "Dark 2") +
  xlab("hexagon id (re-ordered)") + 
  ylab("bin count") +
  ggtitle("(a)") +
  theme(aspect.ratio = 0.75,
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12, angle = 90),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(size = 12))
```

@@

mse_scurve_lwd <- ggplot(error_rm_scurve, 
                     aes(x = benchmark_rm_lwd, 
                         y = log(MSE))) + 
  geom_point(
    size = 1
    ) +
  geom_line(
    linewidth = 0.3
    ) + 
  geom_vline(xintercept = benchmark1, linetype="solid", 
             color = "black", linewidth=0.5, alpha = 0.5) +
  scale_x_continuous("standardized bin count", 
         transform = "reverse") + 
  ylab("log(MSE)") +
  ggtitle("(c)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12, 
            angle=90),
        plot.title = element_text(size = 12))
```

>>>>

## To obtain the standardise counts within hexbins
counts_df1 <- hb_obj_scurve1$std_cts
df_bin_centroids1 <- extract_hexbin_centroids(
  centroids_df = all_centroids_df1, 
  counts_df = counts_df1) |>
  filter(drop_empty == FALSE) |>
  mutate(b1 = "b1 = 7")

@@

## To obtain the standardise counts within hexbins
counts_df2 <- hb_obj_scurve2$std_cts
df_bin_centroids2 <- extract_hexbin_centroids(
  centroids_df = all_centroids_df2, 
  counts_df = counts_df2) |>
  filter(drop_empty == FALSE) |>
  mutate(b1 = "b1 = 11")

@@

## To obtain the standardise counts within hexbins
counts_df3 <- hb_obj_scurve3$std_cts
df_bin_centroids3 <- extract_hexbin_centroids(
  centroids_df = all_centroids_df3, 
  counts_df = counts_df3) |>
  filter(drop_empty == FALSE) |>
  mutate(b1 = "b1 = 17")

@@

The default number of bins $b=b_1\times b_2$ is computed based on the sample size, by setting $b_1=n^{1/3}$, consistent with the Diaconis-Freedman rule [@freedman1981]. The value of $b_2$ is determined analytically by $b_1, q, r_2$. Values of $b_1$ between $2$ and $b_1 = \frac{n}{2}$ are allowed. @fig-param-scurve (b) shows the effect of different choices of $b_1$ on the MSE of the fitted model.

@@

<!--To generate errors for different total number of bins-->
```{r}
#| label: errors-scurve
## To initialize number of bins along the x-axis
bin1_vec_scurve <- 2:15 #sqrt(NROW(training_data_scurve)/2)
#buffer_vec <- seq(0.05, 0.1, by = 0.01)

@@

for (xbins in bin1_vec_scurve) {
  #for (buff in buffer_vec) {

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.07)$bin2

  scurve_model <- fit_highd_model(
    training_data = training_data_scurve,
    emb_df = umap_scurve_scaled,
    bin1 = xbins,
    r2 = r2,
    q = 0.07, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "x"
  )

@@

  ## Compute error
  error_df <- glance(
    df_bin_centroids = df_bin_centroids_scurve,
    df_bin = df_bin_scurve,
    training_data = training_data_scurve,
    newdata = NULL,
    type_NLDR = "UMAP",
    col_start = "x") |>
    mutate(bin1 = xbins,
           bin2 = bin2,
           b = bin1 * bin2,
           b_non_empty = NROW(df_bin_centroids_scurve))

@@

  #}

@@

mse_scurve_b <- ggplot(error_scurve, 
                     aes(x = bin1, 
                         y = MSE)) + 
  geom_point(size = 1) +
  geom_line(linewidth = 0.05) + 
  geom_vline(xintercept = 11, linetype="solid",
             color = "#bdbdbd", linewidth=1) +
  geom_vline(xintercept = 9, linetype=2,
             color = "#bdbdbd", linewidth=1) +
  geom_vline(xintercept = 13, linetype=2,
             color = "#bdbdbd", linewidth=1) +
  labs(x = expression(b[1]), y = "MSE") +
  ggtitle("(b)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12, 
            angle=90),
        plot.title = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        legend.key.height = unit(1, 'cm'),
        legend.key.width = unit(1, 'cm'),
        legend.key.size = unit(1, 'cm'))
```

@@

cell_count_scurve <- ggplot(df_bin_centroids_all, 
                            aes(x = reorder(as.factor(hexID), 
                                            -bin_counts), 
                                y = bin_counts,
                                group = as.factor(b1),
                                color = as.factor(b1))) + 
  geom_line() + 
  # geom_hline(yintercept = benchmark1, color = "black", linewidth=0.5, alpha = 0.5) +
  scale_colour_discrete_qualitative(palette = "Dark 3") +
  xlab("hexagon id (re-ordered)") + 
  ylab("bin count") +
  ggtitle("(a)") +
  theme(aspect.ratio = 0.75,
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12, angle = 90),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        legend.key.size = unit(1, 'cm'),
        legend.key.height = unit(1, 'cm'))
```

@@

mse_scurve_lwd <- ggplot(error_rm_scurve, 
                     aes(x = benchmark_rm_lwd, 
                         y = MSE)) + 
  geom_point(
    size = 1
    ) +
  geom_line(
    linewidth = 0.3
    ) + 
  # geom_vline(xintercept = benchmark1, linetype="solid", 
  #            color = "#bdbdbd", linewidth=1, alpha = 0.5) +
  scale_x_continuous("standardized bin count", 
         transform = "reverse") + 
  ylab("MSE") +
  ggtitle("(c)") +
  theme(aspect.ratio = 0.75,
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12, 
            angle=90),
        plot.title = element_text(size = 12))
```
